The role of advection versus diffusion in intraventricular transport

### 2025-03-18 paper/main.tex
draft abstract, update intro

Cerebrospinal fluid (CSF) flow is integral to brain function.
The underlying mechanisms of cilia-mediated flow and transport
in the brain ventricles remain to be fully explored.
In this paper, we analyze and evaluate the role of motile cilia
in intraventricular CSF flow and transport. We combine computational
modeling with in vivo experiments that monitor transport of a photoconvertible
protein secreted in the brain ventricles.
We use the geometry of the zebrafish larval brain ventricles,
for which we have detailed knowledge of cilia properties and CSF motion velocities and frequencies,
to develop a finite element model.
Our results show that while cilia contribute to advection of large particles,
diffusion plays a significant role in the transport of small molecules such as proteins.
We also found that cilia location and the geometry of the ventricular system impact solute distribution.
Altogether, our work offers both a computational framework that can be applied to other
ventricular system and new concepts of how molecules are transported within the brain and its ventricles.

@@

% I1: What is the problem - from big picture to our precise topic - why is it interesting?
Brain development and function depend on numerous regulatory
processes, including the flow of cerebrospinal fluid
(CSF)~\cite{CITATTION}. Produced mostly by the choroid
plexus and partly by brain interstitial fluid,
CSF is a clear liquid that flows through the ventricular system and subarachnoid
space~\cite{Damkier2013, Brodal2016}. By transporting growth factors, nutrients, neurotransmitters
and metabolites, as well as removing waste products, CSF regulates
brain development and homeostasis~\cite{Ringers2020Role, fame2020emergence, del2010ependymal}.
Disruption of the CSF circulation is detrimental to brain health and disturbs
cognitive and motor functions~\cite{Johanson2008}.
Pathological CSF flow may also lead to
disorders such as hydrocephalus, where accumulation of
CSF often requires emergency surgical
treatment~\cite{duy2022rethinking, kahle2024paediatric, wallmeier2022role},
but for which there is little agreement on optimal treatments~\cite{flannery2014pediatric}. 
Of particular importance is the fact that
hydrocephalus is observed in animal models and humans with deficient motile
cilia~\cite{Afzelius2004Cilia-relatedDiseases, Olstad2019CiliaryDevelopment, Mitchison2017Motile,
Ibanez-Tallon2002LossHydrocephalus},
indicating that the cilia play a crucial role for healthy brain function.

@@

% I3: What do we not know about the problem and what do we want to
% know? What are the precise research questions we want to address?
% Refer to specific studies and what specifically we do not know.

%
%Until now, most studies have analyzed one physical mechanism at a
%time, especially in mammals.
%
%  Studies on animal models such as mice and
% zebrafish have provided unequivocal evidence that cilia are crucial
% for CSF flow in these animals~\cite{Olstad2019CiliaryDevelopment, Faubel2016Cilia}.
% It is clear that cilia affects advective intraventricular transport 
% to some extent. However a quantification of this effect, as well as of to which
% extent diffusion plays a role in transporting solutes, remains
% yet to be performed.

To date, it remains poorly understood how motile cilia contribute to 
the overall fluid flow and solute transport within the brain.
Experimental endeavors are challenging to pursue due to
technical limitations: the microscopic size and relatively 
high beating frequencies of the cilia~\cite{ocallaghan2012analysis, mahuzier2018ependymal,
Jeong2022MeasurementTelencephalon, Olstad2019CiliaryDevelopment, spassky2013motile, DGama2021Diversity},
as compared to other fluid dynamic processes in the brain that are slower and span larger
spatial scales~\cite{Vinje2019RespiratoryMeasurements, eide2024functional, fultz2019coupled, mestre2018flow},
make cilia difficult to monitor in vivo in humans.
Until now, most studies have analyzed one physical mechanism at a time, especially in mammals.
For instance, studies using brain explants have convincingly shown
that cilia generate complex flow patterns at the surface of the brain
ventricles~\cite{Faubel2016Cilia-basedVentricles, Sawamoto2006NewBrain, ibanez2003beat};
however, they failed to identify the role of pressure
gradients associated with respiration and cardiac pulsations in CSF
flow and the accompanying transport. Similarly, while MRI technology
can measure CSF movement in humans~\cite{feinberg1987human, battal2011cerebrospinal,
yamada2015current}, it cannot detect the microscopic
movements of cilia to understand their impact in the large human
brain~\cite{Siyahhan2014FlowVentricles}. Consequently, we lack a comprehensive understanding of how
cilia contribute to CSF motion and solute transport, and whether cilia
dysfunction alters CSF composition and thus leads to brain disorders.

The technical limitations of studying cilia-mediated flow and transport opens the
possibility of mathematical modeling and computational investigations. Numerous previous studies
have focused on modeling the interaction between single or several cilia and surrounding
fluids~\cite{Guo2020SimulatingGeometries, Ruvalcaba2021NumericalTree, Smith2009MathematicalFluids,
Cui2019NumericalMethod, Cui2022AFlow}.
Others considered applications such as ciliary transport of mucous and particles in
airways~\cite{Fulford1986Muco-ciliaryLung, Ramirez-SanJuan2020Multi-scaleArrays, roth2025structure},
propulsion and locomotion of ciliated organisms~\cite{BLAKE1974MechanicsMotion,
Jahn1972LocomotionProtozoa, Brennen1977FluidFlagella},
oxygen regulation in coral reefs~\cite{Pacherres2022CiliaryProduction},
and cilia-mediated flow in the brain~\cite{Siyahhan2014FlowVentricles,
Yoshida2022EffectVentricles, Salman2022ComputationalEmbryo, Thouvenin2020OriginCanal}.
These simulation studies have provided valuable insights on how cilia mediate complex
biological flows, and how cilia may contribute to tissue physiology.

% I4: How do we address the problem and what do we find out?
In this study, we combined computational and experimental approaches to investigate
the role of motile cilia in intraventricular CSF flow and transport.
First, we computed flow patterns and simulated solute transport with a computational model. To
this end, we used the geometry of the zebrafish larval brain ventricles,
for which we have detailed knowledge of cilia properties and CSF motion velocities
and frequencies. Notably, this early developmental stage allows us to simplify our model
by omitting CSF secretion, which starts at later stages~\cite{Jeong2024TheZebrafish},
while still being well-conserved with mammals~\cite{DGama2025MotileBrain,
Olstad2019CiliaryDevelopment, Ringers2020Role, DGama2021Diversity, jurisch2020radial}. 
Next, we validated our computation model with experimental data by monitoring in vivo
the dynamics of a photoconvertible fluorescent protein Dendra2
expressed within the brain ventricles.
Our results show that diffusion plays a major role in the transport of small solutes,
whereas cilia majorly contribute to the movement of larger particles.
We also found that the location of cilia, as well as the ventricular geometry, largely
impact solute distribution. Altogether, our work presents a computational
framework together with new concepts of how cilia distribute molecules and
particles within the brain ventricles.

@@

% Why? (Intersting point #1. However, we do not know aspect #2. 
% How? (We therefore ...)
% What ...

In-vivo particle tracking reveals characteristic CSF flow patterns in
zebrafish larval brain ventricles
(\Crefpanels{fig:fig1}{a--c})~\cite{Olstad2019CiliaryDevelopment}. We
first ask whether we can accurately represent these flow patterns with
computational fluid dynamics driven by cardiac pulsations and motile
cilia lining the brain ventricles.

@@

\begin{figure}
    \centering
    \includegraphics[width=\linewidth]{graphics/figure1.png}
    \caption{
    \textbf{a.} Schematic illustration of the two cerebrospinal fluid (CSF) flow components modeled,
    motile cilia and cardiac pulsations, and their contributions to the flow patterns.
    \textbf{b, c.} Particle tracking data visualizing CSF flow directionality (\textbf{b}) and
    velocity magnitudes in the middle ventricle.
    \textbf{d.} Computational mesh with marked facets indicating regions of motile cilia (magenta) and
    anterior/posterior facets (blue).
    \textbf{e.} Streamlines of the (steady-state) CSF flow field simulated with the
    cilia-only/no-cardiac flow model.
    \textbf{f.} Streamlines of the CSF flow field simulated with the cardiac-only/no-cilia flow
    model at a time 25\% into the cardiac cycle. %t=0.112613 s, period=0.45045 s
    \textbf{g.} Illustration of the cardiac-related normal pressure forcing imposed on the
    anterior/posterior boundaries.
    \textbf{h--j.} Streamlines simulated with the baseline flow model at
    a time a time 25\% (\textbf{h}), 50\% (\textbf{i}) and 75\% (\textbf{j})
    into the cardiac cycle. Times indicated in panel \textbf{g}.
    %t=0.112613 s %t=0.225225 s %t=0.315315 s
    Scale bars 50 \textmu m.
    }
    \label{fig:fig1}
\end{figure}

>>>>

Cerebrospinal fluid (CSF) is integral to brain function, as the fluid
provides mechanical support of the brain and the flow of CSF helps
distribute nutrients, neurotransmitters and metabolites throughout
the central nervous system. CSF flow is partly driven
by beating of motile cilia, hair-like structures that line the ependymal cell layer
of the brain ventricles. Despite the physiological importance of CSF flow,
the underlying mechanisms of cilia-mediated flow and transport
in the brain ventricles remain yet to be comprehensively resolved.
In this study, we analyze and evaluate the role of motile cilia
in intraventricular CSF flow and transport. We use the geometry
of the zebrafish larval brain ventricles, for which we have detailed knowledge of cilia properties and CSF motion velocities
and frequencies, to develop finite element methods that model flow and transport.
The computational model is validated by in vivo experiments that monitor transport
of a photoconvertible protein secreted in the brain ventricles.
Our results show that while cilia contribute to advection of large particles,
diffusion plays a significant role in the transport of small molecules such as proteins.
We also found that cilia location and the geometry of the ventricular system impact solute distribution.
Altogether, our work presents a computational framework that can be applied to other
ventricular systems, together with new concepts of how molecules are transported within the brain and its ventricles.

@@

% I1: What is the problem - from big picture to our precise topic - why is it interesting?
Brain development and function depend on numerous regulatory
processes, including the flow of cerebrospinal fluid
(CSF)~\cite{fame2020emergence}. Produced mostly by the choroid
plexus and partly by brain interstitial fluid~\cite{Brodal2016},
CSF is a clear liquid that flows through the ventricular system and subarachnoid
space~\cite{Damkier2013}. By transporting growth factors, nutrients, neurotransmitters
and metabolites, as well as removing waste products, CSF regulates
brain development and homeostasis~\cite{Ringers2020Role, fame2020emergence, del2010ependymal}.
Disruption of the CSF circulation is detrimental to brain health and disturbs
cognitive and motor functions~\cite{Johanson2008}.
Pathological CSF flow may also lead to
disorders such as hydrocephalus, where accumulation of
CSF often requires emergency surgical
treatment~\cite{duy2022rethinking, kahle2024paediatric, wallmeier2022role},
but for which there is little agreement on optimal treatments~\cite{flannery2014pediatric}.

@@

% I3: What do we not know about the problem and what do we want to
% know? What are the precise research questions we want to address?
% Refer to specific studies and what specifically we do not know.
To date, it remains poorly understood how motile cilia contribute to 
the overall fluid flow and solute transport within the brain.
Experimental endeavors are challenging to pursue due to
technical limitations: the microscopic size and relatively 
high beating frequencies of the cilia~\cite{ocallaghan2012analysis, mahuzier2018ependymal,
Jeong2022MeasurementTelencephalon, Olstad2019CiliaryDevelopment, spassky2013motile, DGama2021Diversity},
as compared to other fluid dynamic processes in the brain that are slower and span larger
spatial scales~\cite{Vinje2019RespiratoryMeasurements, eide2024functional, fultz2019coupled, mestre2018flow},
make cilia difficult to monitor in vivo in humans.
Until now, most studies have analyzed one physical mechanism at a time, especially in mammals.
For instance, studies using brain explants have convincingly shown
that cilia generate complex flow patterns at the surface of the brain
ventricles~\cite{Faubel2016Cilia-basedVentricles, Sawamoto2006NewBrain, ibanez2003beat};
however, they failed to identify the role of other contributions to CSF
flow and accompanying transport, such as pressure
gradients associated with respiration and cardiac pulsations. Moreover, while MRI technology
can measure CSF movement in humans~\cite{feinberg1987human, battal2011cerebrospinal,
yamada2015current}, it cannot detect the microscopic
movements of cilia to understand their impact in the large human
brain~\cite{Siyahhan2014FlowVentricles}. 
These technical limitations of studying cilia-mediated flow and transport make 
mathematical modeling and computational investigations an attractive alternative to experimental
studies. Numerous previous simulation studies have focused on modeling the interaction
between single or several cilia and surrounding
fluids~\cite{Guo2020SimulatingGeometries, Ruvalcaba2021NumericalTree, Smith2009MathematicalFluids,
Cui2019NumericalMethod, Cui2022AFlow}.
Others considered applications such as ciliary transport of mucous and particles in
airways~\cite{Fulford1986Muco-ciliaryLung, Ramirez-SanJuan2020Multi-scaleArrays, roth2025structure},
propulsion and locomotion of ciliated organisms~\cite{BLAKE1974MechanicsMotion,
Jahn1972LocomotionProtozoa, Brennen1977FluidFlagella},
oxygen regulation in coral reefs~\cite{Pacherres2022CiliaryProduction},
and cilia-mediated flow in the brain~\cite{Siyahhan2014FlowVentricles,
Yoshida2022EffectVentricles, Salman2022ComputationalEmbryo, Thouvenin2020OriginCanal}.
These simulation studies have provided valuable insights on how cilia mediate complex
biological flows, and how cilia may contribute to tissue physiology.
However, we still lack a comprehensive understanding of how
cilia contribute to CSF motion and solute transport within the brain.
This is of particular importance since disorders such as hydrocephalus are observed
in animal models and humans with deficient motile cilia~\cite{Afzelius2004Cilia-relatedDiseases,
Olstad2019CiliaryDevelopment, Mitchison2017Motile, Ibanez-Tallon2002LossHydrocephalus},
indicating that the cilia play a crucial role for healthy brain function.

% I4: How do we address the problem and what do we find out?
In this study, we combined computational and experimental approaches to investigate
the role of motile cilia in intraventricular CSF flow and transport.
We model the flow and transport by partial differential equations and solve
them numerically with finite element methods to compute flow patterns and
solute concentration dynamics. A geometry representing the zebrafish larval brain ventricles
was used, for which we have detailed knowledge of cilia properties and CSF motion velocities
and frequencies~\cite{Olstad2019CiliaryDevelopment}. 
Notably, this early developmental stage allows us to simplify our model
by omitting CSF secretion, which starts at later stages~\cite{Jeong2024TheZebrafish},
while still being well-conserved with mammals~\cite{DGama2025MotileBrain,
Olstad2019CiliaryDevelopment, Ringers2020Role, DGama2021Diversity, jurisch2020radial}. 
The computational model is validated with experimental data by monitoring in vivo
the dynamics of a photoconvertible fluorescent protein Dendra2
expressed within the brain ventricles.
Our results show that diffusion plays a major role in the transport of small solutes,
whereas cilia majorly contribute to the movement of larger particles.
We also found that the location of cilia, as well as changes to the ventricular geometry,
largely impact solute distribution. In summary, our work presents a computational
framework together that can be applied to other
ventricular systems in the future, together with new concepts
of how cilia distribute molecules and
particles within the brain ventricles.

@@

% Why? (Interesting point #1. However, we do not know aspect #2. 
% How? (We therefore ...)
% What ...

In vivo particle tracking reveals characteristic CSF flow patterns in
zebrafish larval brain ventricles
(\Crefpanels{fig:fig1}{a--c})~\cite{Olstad2019CiliaryDevelopment}. We
first ask whether we can accurately represent these flow patterns with
computational fluid dynamics driven by cardiac pulsations and motile
cilia lining the surface of the brain ventricles.

@@

\begin{figure}
    \centering
    \includegraphics[width=\linewidth]{graphics/figure1.png}
    \caption{
    \textbf{a.} Schematic illustration of the two cerebrospinal fluid (CSF) flow components modeled,
    motile cilia and cardiac pulsations, and their contributions to the flow patterns.
    \textbf{b, c.} Particle tracking data visualizing CSF flow directionality (\textbf{b}) and
    velocity magnitudes (\textbf{c}) in the middle ventricle.
    \textbf{d.} Computational mesh with marked facets indicating regions of motile cilia (magenta) and
    anterior/posterior facets (blue).
    \textbf{e.} Streamlines of the (steady-state) CSF flow field simulated with the
    cilia-only/no-cardiac flow model.
    \textbf{f.} Streamlines of the CSF flow field simulated with the cardiac-only/no-cilia flow
    model at a time 25\% into the cardiac cycle. %t=0.112613 s, period=0.45045 s
    \textbf{g.} Illustration of the cardiac-related normal pressure forcing imposed on the
    anterior/posterior boundaries.
    \textbf{h--j.} Streamlines simulated with the baseline flow model at
    a time a time 25\% (\textbf{h}), 50\% (\textbf{i}) and 75\% (\textbf{j})
    into the cardiac cycle. Times indicated in panel \textbf{g}.
    %t=0.112613 s %t=0.225225 s %t=0.315315 s
    Scale bars 50 \textmu m.
    }
    \label{fig:fig1}
\end{figure}
